---
title: Understanding the recent increase in black infant mortality in the US
output: word_document
---

Corinne A Riddell, PhD

Sam Harper, PhD

Jay S Kaufman, PhD

Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics &
Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: corinne.riddell@mail.mcgill.ca

Current word count (text only, exclusive of title, references, and tables): 596

Date of last revision: January 16, 2017

**Introduction**

2015 witnessed the first annual decrease in life expectancy in the US since 1993.^1^ While life expectancy decreased for both whites and blacks, black men fared the worst, losing nearly twice as many years of life expectancy as white men.^2^ It was also observed that the infant mortality rate increased for blacks while decreasing for whites between 2014 and 2015. We investigated trends in infant deaths from specific causes for blacks and whites over the past ten years in the US, and report if any specific causes of death increased concomitantly with the recent increase in infant mortality among blacks. 

**Methods**

We abstracted data on deaths and population for Non-Hispanic black and white infants from the US National Vital Statistics System by cause of infant death for 2005-2015.^3^ The infant mortality rate was calculated as the number of deaths divided by the population size, according to race, year, and cause of death. We calculated rates for the top four causes of infant death and for a residual category encapsulating all other causes. The excess event rate among blacks was calculated as the absolute difference between the rates. We do not include confidence intervals because all deaths are included and therefore are not subject to sampling error.^4^ R version 3.2.4 was used to conduct this analysis. No ethical review was required since this study used de-identified data. 

```{r load_libraries, echo=F, warning=F, message=F}
library(ggplot2)
library(plotly)
library(dplyr)
library(gridExtra)
library(grid)
library(epiR)
```

```{r import_cod_data, echo=F}
#read in the CDC wonder COD-specific mortality data
#for NH blacks and NH whites, stratified by year (1999-2015)
im.cod.dat <-  read.csv("/Users/corinneriddell/Documents/repos/InfantMortality/Data/InfantMortality_by_COD_cleaned.csv", 
                        header = T)

im.cod.dat <- im.cod.dat %>% 
  filter(Race %in% c("Black or African American", "White"),
         Hispanic_Origin == "Not Hispanic or Latino") %>%
  select(Race, Year, Deaths, Population, ICD10.130.Infants, Crude_Rate) %>%
  mutate(Race = relevel(Race, ref = "White")) %>%
  mutate(Race = plyr::revalue(Race, c("White"="White", "Black or African American"="Black")),
         Population = as.numeric(as.character(Population)),
         RatePer1000 = (Deaths/Population)*1000,
         ICD10 = as.character(ICD10.130.Infants))
```

```{r create_cod_aggregated_variables, echo=F}
#cod_recode marks those rows that correspond to the FOURTEEN leading causes of death
im.cod.dat$cod_recode <- NA
im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == 
                        "#Congenital malformations, deformations and chromosomal abnormalities (Q00-Q99)"] <- "Congenital Malformations"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Congenital Malformations" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == 
                        "#Disorders related to short gestation and low birth weight, not elsewhere classified (P07)"] <- "Short Gestation/Low Birthweight"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Short Gestation/LBW" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Sudden infant death syndrome (R95)"] <- "SIDS"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "SIDS" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Newborn affected by maternal complications of pregnancy (P01)"] <- "Maternal Complications"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Maternal Complication of Pregnancy" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Accidents (unintentional injuries) (V01-X59)"] <- "Accidents"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Accidents" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Newborn affected by complications of placenta, cord and membranes (P02)"] <- "Complications of Placenta, Cord, and Membranes"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Complications of Placenta, Cord, and Membranes" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Respiratory distress of newborn (P22)"] <- "Respiratory Distress"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Respiratory Distress" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Bacterial sepsis of newborn (P36)"] <- "Bacterial Sepsis"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Bacterial Sepsis" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Diseases of the circulatory system (I00-I99)"] <- "Diseases of the Circulatory System"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Diseases of the Circulatory System" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Neonatal hemorrhage (P50-P52,P54)"] <- "Neonatal Hemorrhage"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Neonatal Hemorrhage" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Intrauterine hypoxia and birth asphyxia (P20-P21)"] <- "Intrauterine hypoxia and birth asphyxia"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Intrauterine hypoxia and birth asphyxia" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Necrotizing enterocolitis of newborn (P77)"] <- "Necrotizing enterocolitis"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Necrotizing enterocolitis" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Atelectasis (P28.0-P28.1)"] <- "Atelectasis"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Atelectasis" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Assault (homicide) (*U01,X85-Y09)"] <- "Assault"
#table(im.cod.dat$cod_recode)
#im.cod.dat[im.cod.dat$cod_recode == "Assault" & is.na(im.cod.dat$cod_recode) == F, ]

#remove other rows of data
im.cod.dat2 <- im.cod.dat %>% filter(is.na(cod_recode) == F) %>%
  select(Race, Year, Deaths, Population, RatePer1000, cod_recode)

rm(im.cod.dat)

#calculate the infant mortality rate for the top FOUR causes of infant death, by year and race
total4 <- im.cod.dat2 %>% group_by(Race, Year) %>%
  filter(cod_recode %in% c("Congenital Malformations", "SIDS", 
                           "Short Gestation/Low Birthweight", "Maternal Complications")) %>%
  summarise(Deaths.top4 = sum(Deaths))

im.cod.dat_top4 <- im.cod.dat2 %>% 
  filter(cod_recode %in% c("Congenital Malformations", "SIDS", 
                           "Short Gestation/Low Birthweight", "Maternal Complications")
         )
```

```{r setup_data, echo=F}
#read in a separate file, where the deaths are aggregated over all COD
#note: this dataset is stratified by gender, but the previous one was not
im.dat <-  read.csv("/Users/corinneriddell/Documents/repos/InfantMortality/Data/InfantMortalityData.csv",
                      header = T)

imdat.subset <- im.dat %>% 
  filter(Race %in% c("Black or African American", "White"),
         Hispanic_Origin == "Not Hispanic or Latino") %>%
  select(Race, Year, Gender, Deaths, Population, Crude_Rate) %>%
  mutate(Race = relevel(Race, ref = "White")) %>%
  mutate(Race = droplevels(Race)) %>%
  mutate(Race = plyr::revalue(Race, c("White"="White", "Black or African American"="Black")),
         Population = as.numeric(as.character(Population)),
         RatePer1000 = (Deaths/Population)*1000)

rm(im.dat)

#calculate the total infant deaths by race and year
totals <- imdat.subset %>% group_by(Race, Year) %>% 
  summarise(Deaths.all = sum(Deaths), #this sums across gender
            Population = sum(Population)) 

totals_top4 <- merge(totals, total4, by = c("Race", "Year"))

rm(totals, total4)

#calculate the rate of deaths from all other causes
totals_top4 <- totals_top4 %>% mutate(Deaths = Deaths.all - Deaths.top4,
                                      RatePer1000 = (Deaths/Population)*1000,
                                      cod_recode = "All other causes") %>%
                               select(Race, Year, Population, Deaths, RatePer1000, cod_recode) 

#add in the "All other causes" data to the COD-specific data
im.cod.dat_top4 <- rbind(im.cod.dat_top4, totals_top4)  
rm(totals_top4)

#order the COD variable 
im.cod.dat_top4$cod_recode <- factor(im.cod.dat_top4$cod_recode)
im.cod.dat_top4$cod_recode <- reorder(im.cod.dat_top4$cod_recode, im.cod.dat_top4$RatePer1000, max)
im.cod.dat_top4$cod_recode2 <- factor(im.cod.dat_top4$cod_recode, levels = rev(levels(im.cod.dat_top4$cod_recode)))
```

```{r reshape_long_to_wide, echo=F}
black2 <- subset(im.cod.dat_top4, Race == "Black") %>% 
  select(Year, RatePer1000, Population, Deaths, cod_recode, cod_recode2) %>%
  rename(Black_IMR = RatePer1000, Black.Pop = Population,
         Black.Deaths = Deaths)

white2 <- subset(im.cod.dat_top4, Race == "White") %>% 
  select(Year, RatePer1000, Population, Deaths, cod_recode, cod_recode2) %>%
  rename(White_IMR = RatePer1000, White.Pop = Population,
         White.Deaths = Deaths)

both.cod <- merge(black2, white2, by = c("Year", "cod_recode", "cod_recode2"))
rm(black2, white2)

#add risk ratio and risk difference to the data frame
both.cod <- both.cod %>% mutate(RR = Black_IMR/White_IMR, RD = Black_IMR - White_IMR)

#calculate for overall rates (not COD-specific)
both <- both.cod %>% group_by(Year) %>% 
                     summarise(Black.Deaths = sum(Black.Deaths), 
                               White.Deaths = sum(White.Deaths),
                               Black.Pop = first(Black.Pop),
                               White.Pop = first(White.Pop)) %>%
                     mutate(Black_IMR = (Black.Deaths/Black.Pop)*1000,
                            White_IMR = (White.Deaths/White.Pop)*1000,
                            RR = Black_IMR/White_IMR, 
                            RD = Black_IMR - White_IMR)
                     
```

**Results**

From 2005 to 2012, the infant mortality rate for Non-Hispanic blacks decreased from `r round(both$Black_IMR[both$Year == 2005], 1)` to `r round(both$Black_IMR[both$Year == 2012], 1)` per 1,000 infants (**Figure 1a**). Thereafter, the infant mortality rate in blacks plateaued, and increased from `r formatC(round(both$Black_IMR[both$Year == 2014], 1), format='f', digits = 1)` to `r round(both$Black_IMR[both$Year == 2015], 1)` per 1,000 from 2014 to 2015. For Non-Hispanic whites, the rate decreased monotonically from `r round(both$White_IMR[both$Year == 2005], 1)` to `r round(both$White_IMR[both$Year == 2015], 1)` per 1,000 from 2005 to 2015. Because the black infant mortality rate declined faster than the white rate, excess events in blacks fell from `r round(both$RD[both$Year == 2005], 1)` deaths per 1,000 infants in 2005 to `r formatC(round(both$RD[both$Year == 2012], 1), format='f', digits = 1)` deaths per 1,000 in 2012, but rose to `r round(both$RD[both$Year == 2015], 1)` in 2015 (**Figure 1b**). These excess events imply nearly `r round(both$Black.Deaths[both$Year == 2015] - (both$White_IMR[both$Year == 2015]/1000*both$Black.Pop[both$Year == 2015]), -3)` additional infant deaths among the `r formatC(both$Black.Pop[both$Year == 2015], format='f', digits=0)` blacks infants compared to the number of deaths that would have occurred had they experienced the mortality rate among whites.

**Figure 2** shows trends in cause-specific mortality for the four leading causes of infant death and all other causes. Between 2005 and 2011, deaths from short gestation and low birthweight decreased for blacks, but have plateaued in recent years. For other leading causes, rates among blacks and whites decreased between 2005 and 2015, although deaths related to both Sudden Infant Death Syndrome (SIDS) and congenital malformations increased for blacks last year. Progress has stalled with respect to all other causes of infant death in recent years, and last year also saw an increase for blacks, but not for whites.

**Comment**

The sustained progress in reducing infant mortality among blacks since 2005 stalled in the past few years, leading to increases in the absolute inequality in infant mortality between blacks and whites in two of the past three years. No single cause appears solely responsible for the recent increase in black infant mortality. The preterm birth rate is nearly 50% higher for blacks compared with whites,^5^ and black infants experience three times as many deaths related to short gestation/low birthweight, making it the leading cause of infant death among blacks. Interventions to further reduce the rate of preterm birth among blacks appear the most promising option for reducing black infant mortality, and the absolute inequality between black and white infants. 

To confirm our findings, linked birth/death files should be used once they become available for the most recent years.

**Author Contributions**

Dr. Riddell had full access to all of the data and takes responsibility for the integrity of the data and the accuracy of the analysis. A replication data set including the raw data and statistical code to reproduce the paper in its entirety is publicly available on github: https://github.com/corinne-riddell/InfantMortality/.

Study concept and design: Riddell, Harper, Kaufman

Acquisition of data: Riddell 

Analysis and interpretation of data: Riddell, Harper, Kaufman 

Drafting of the manuscript: Riddell, Harper, Kaufman

Critical revision of the manuscript for important intellectual content: Riddell, Harper, Kaufman

Statistical analysis: Riddell 

Obtained funding: No funding obtained 

Administrative, technical, or material support: None 

Study supervision: Riddell, Harper, Kaufman

Acknowledgements: Corinne Riddell is a post-doctoral researcher at McGill University. Sam Harper was supported by a Chercheur Boursier Junior 1 from the Fonds de la Recherche en Santé du Québec. Jay Kaufman was supported by a Canada Research Chair in Health Disparities. The sources of support had no control over the design and conduct of the study; collection, management, analysis, and interpretation of the data; and preparation, review, or approval of the manuscript. 


**References**

1. Xu J, Murphy SL, Kochanek KD, Arias E. Mortality in the United States, 2015. NCHS Data Brief. 2016;(267):1–8.

2. Harper S, Kaufman JS, Cooper RS. Declining United States Life Expectancy: A First Look. Submitted to JAMA.

3. Centers for Disease Control and Prevention, National Center for Health Statistics. Underlying cause of death 1999-2015 on CDC WONDER online database, released December, 2016. Data are from the Multiple Cause of Death Files, 1999-2015, as compiled from data provided by the 57 vital statistics jurisdictions through the Vital Statistics Cooperative Program; December 2016. Available from: https://wonder.cdc.gov/ [cited 22 Dec 2016]. 

4. Kochanek KD, Murphy SL, Xu JQ, Tejada-Vera B. Deaths: Final data for 2014. National vital statistics reports; vol 65 no 4. Hyattsville MD: National Center for Health Statistics. 2016.

5. March of Dimes. 2016 Premature birth report card. http://www.marchofdimes.org/materials/premature-birth-report-card-united-states.pdf. Accessed on Dec 31, 2016.

**Figure Legends**

**Figure 1: Infant mortality and excess risk of death for Non-Hispanic blacks compared with whites, United States, 2005-2015**

```{r imr_plot, echo=F}
imr_plot_1 <- ggplot(subset(both, Year >= 2005), aes(x = Year, y = Black_IMR))  + 
  geom_segment(aes(y = 0, yend = 14.3, x=2005, xend=2005), lty =3, col = "grey37", lwd = 0.5) +
  geom_segment(aes(y = 0, yend = 11.1, x=2012, xend=2012), lty =3, col = "grey37", lwd = 0.5) + 
  geom_segment(aes(y = 0, yend = 11.2, x=2015, xend=2015), lty =3, col = "grey37", lwd = 0.5) +
  geom_line(aes(col = "Black")) + 
  geom_line(aes(y = White_IMR, col = "White")) +
  ylab("Rate per 1,000 infants") + 
  scale_y_continuous(limits = c(0, 15.5), breaks = seq(from=0, to=15, by = 3), minor_breaks = seq(from=0, to=15, by = 1)) + 
  scale_x_continuous(limits = c(2005, 2017), breaks = seq(from=2005, to=2015.5, by = 2)) +
  scale_color_manual(values = c("#b2182b", "#2166ac"), guide = guide_legend(title = "")) +
  theme_bw() + theme(legend.position = "none") + ggtitle("a) Infant Mortality Rate") +
  geom_text(aes(x = 2016, y = 4.9), label = "White", col = "#2166ac", check_overlap = T) +
  geom_text(aes(x = 2016, y = 11.2), label = "Black", col = "#b2182b", check_overlap = T) +
  geom_text(data = subset(both, Year %in% c(2005, 2012, 2015)), 
            aes(x = Year, y = Black_IMR + 1, label = round(Black_IMR,1)), col = "#b2182b") +
  geom_text(data = subset(both, Year %in% c(2005, 2012, 2015)), 
            aes(x = Year, y = White_IMR + 1, label = round(White_IMR,1)), col = "#2166ac")  

```

```{r RD_plot, echo=F}
imr_plot_2 <- ggplot(subset(both, Year >= 2005), aes(x = Year, y = RD))  + 
  geom_segment(aes(y = 0, yend = 8.4, x = 2005, xend = 2005), lty = 3, col = "grey37", lwd = 0.5) +
  geom_segment(aes(y = 0, yend = 6.0, x = 2012, xend = 2012), lty = 3, col = "grey37", lwd = 0.5) + 
  geom_segment(aes(y = 0, yend = 6.3, x = 2015, xend = 2015), lty = 3, col = "grey37", lwd = 0.5) +
  geom_line() + 
  ylab("Excess events per 1,000 infants") + 
  scale_y_continuous(limits = c(0, 10), breaks = seq(from=0, to=10, by = 3), minor_breaks = seq(from=0, to=9, by = 1)) +
  scale_x_continuous(limits = c(2005, 2017), breaks = seq(from=2005, to=2015.5, by = 2)) +
  theme_bw() + ggtitle("b) Excess events") +
  geom_text(data = subset(both, Year %in% c(2005, 2012, 2015)), 
            aes(x = Year, y = RD + 0.5, label = round(RD,1))) 
```

```{r figure1, fig.width=7, fig.height=3.5, echo=F, eval = T}
grid.arrange(imr_plot_1, imr_plot_2, ncol = 2)
ggsave(file = "/Users/corinneriddell/Documents/repos/InfantMortality/Figure1.tiff", 
       arrangeGrob(imr_plot_1, imr_plot_2, ncol = 2),
       width = 7, height =3)
```

```{r figure2_setup, echo=F, warning=F, eval=T}
a <- ggplot(subset(both.cod, Year >= 2005 & cod_recode!= "All other causes"), aes(x = Year, y = Black_IMR*100)) + 
            geom_segment(data = subset(both.cod, Year %in% c(2005, 2012, 2015) & cod_recode!= "All other causes"), 
                         aes(yend = Black_IMR*100, y = 0, xend = Year), lty = 3) +
            geom_line(aes(col = "Black")) + facet_wrap( ~ cod_recode2) +
            geom_line(aes(y = White_IMR*100, col = "White")) +   
            theme_bw() + ylab("Rate per 100,000 infants") + 
            theme(legend.position = "top") +
            scale_color_manual(values = c("#b2182b", "#2166ac"), guide = guide_legend(title = "")) +
            scale_y_continuous(limits = c(-10, 4*100), breaks = seq(from=0, to=4*100, by = 1*100), 
                               minor_breaks = seq(from=0, to=4*100, by = 0.5*100)) +
            scale_x_continuous(limits = c(2004, 2016), breaks = seq(from=2005, to=2015.5, by = 2)) +
            geom_text(data = subset(both.cod, Year %in% c(2005, 2012, 2015) & cod_recode!= "All other causes"), 
                      aes(x = Year, y = (Black_IMR + 0.3)*100, label = round(Black_IMR*100,1)), col = "#b2182b") +
            geom_text(data = subset(both.cod, Year %in% c(2005, 2012, 2015) & cod_recode!= "All other causes"), 
                      aes(x = Year, y = (White_IMR - 0.3)*100, label = round(White_IMR*100,1)), col = "#2166ac") 

b <- ggplot(subset(both.cod, Year >= 2005 & cod_recode== "All other causes"), aes(x = Year, y = Black_IMR*100)) + 
            geom_segment(data = subset(both.cod, Year %in% c(2005, 2012, 2015) & cod_recode== "All other causes"), 
                         aes(yend = Black_IMR*100, y = 0, xend = Year), lty = 3) +
            geom_line(aes(col = "Black")) + facet_wrap(~cod_recode2) +
            geom_line(aes(y = White_IMR*100, col = "White")) +   
            theme_bw() + ylab("Rate per 100,000 infants") +
            scale_color_manual(values = c("#b2182b", "#2166ac"), guide = FALSE) +
            scale_y_continuous(limits = c(-10, 8*100), breaks = seq(from=0, to=8*100, by = 1*100), 
                               minor_breaks = seq(from=0, to=8*100, by = 0.5*100)) +
            scale_x_continuous(limits = c(2004, 2016), breaks = seq(from=2005, to=2015.5, by = 2)) +
            geom_text(data = subset(both.cod, Year %in% c(2005, 2012, 2015) & cod_recode== "All other causes"), 
                      aes(x = Year, y = (Black_IMR + 0.5)*100, label = round(Black_IMR*100,1)), col = "#b2182b") +
            geom_text(data = subset(both.cod, Year %in% c(2005, 2012, 2015) & cod_recode== "All other causes"), 
                      aes(x = Year, y = (White_IMR - 0.5)*100, label = round(White_IMR*100,1)), col = "#2166ac")  

blank <- grid.rect(gp=gpar(col="white"))
```

**Figure 2: Trends in COD-specific mortality for the top four causes of infant death and all other causes combined for Non-Hispanic blacks and whites, United States, 2005-2015**

```{r figure2,  echo=F, fig.width=5.5, fig.height=8.25, warning=F, eval=T}
grid.arrange(a, b, blank, 
             layout_matrix = rbind(c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1), 
                                   c(2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3)), 
             heights = c(0.67, 0.33))

ggsave("/Users/corinneriddell/Documents/repos/InfantMortality/Figure2.tiff", 
       arrangeGrob(a, b, blank, 
                   layout_matrix = rbind(c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1), 
                                         c(2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3)), 
                   heights = c(0.67, 0.33)), height = 8, width = 6.5)
```

**Figure 2 Footnotes:**

*International Classification of Diseases, Tenth Revision* (ICD-10) categories taken from the National Center for Health Statistics list of 130 Causes of infant death: Disorders related to short gestation and low birth weight, not elsewhere classified (P07); Congenital malformations, deformations and chromosomal abnormalities (Q00-Q99); Sudden infant death syndrome (R95); Newborn affected by maternal complications of pregnancy (P01); All other causes (all other codes)

**Appendix**

**15 leading causes of death by ICD code for infants in 2014**
ref: https://www.cdc.gov/nchs/data/dvs/lcwk7_2014.pdf

|Cause of death                                                       | ICD-10-CM codes |
|---------------------------------------------------------------------|-----------------|
|Congenital malformations, deformations and chromosomal abnormalities | Q00-Q99 |
|Disorders related to short gestation and low birth weight, not elsewhere classified | P07 |
|Sudden infant death syndrome                                         | R95 |
|Newborn affected by maternal complications of pregnancy              | P01 |
|Accidents (unintentional injuries)                                   | V01-X59 |
|Newborn affected by complications of placenta, cord and membranes    | P02 |
|Respiratory distress of newborn                                      | P22 |
|Bacterial sepsis of newborn                                          | P36 |
|Diseases of the circulatory system                                   | I00-I99 |
|Neonatal hemorrhage                                                  | P50-P52,P54 |
|Intrauterine hypoxia and birth asphyxia                              | P20-P21 |
|Necrotizing enterocolitis of newborn                                 | P77 |
|Atelectasis                                                          | P28.0-P28.1 |
|Assault (homicide)                                                   | *U01,X85-Y09 |
|Diarrhea and gastroenteritis of infectious origin                    | A09 |


