---
title: "Infant Mortality"
author: "Corinne Riddell"
date: "December 16, 2016"
output: html_document
---

```{r, echo=F, warning=F, message=F}
library(ggplot2)
library(plotly)
library(dplyr)
library(gridExtra)
library(grid)
library(epiR)
```

```{r cod_data}
#read in the CDC wonder COD-specific mortality data
#for NH blacks and NH whites, stratified by year (1999-2015)
im.cod.dat <-  read.csv("/Users/corinneriddell/Documents/repos/InfantMortality/Data/InfantMortality_by_COD_cleaned.csv", header = T)

im.cod.dat <- im.cod.dat %>% 
  filter(Race %in% c("Black or African American", "White"),
         Hispanic_Origin == "Not Hispanic or Latino") %>%
  select(Race, Year, Deaths, Population, ICD10.130.Infants, Crude_Rate) %>%
  mutate(Race = relevel(Race, ref = "White")) %>%
  mutate(Race = plyr::revalue(Race, c("White"="White", "Black or African American"="Black")),
         Population = as.numeric(as.character(Population)),
         RatePer1000 = (Deaths/Population)*1000,
         ICD10 = as.character(ICD10.130.Infants))

#cod_recode marks those rows that correspond to the 14 leading causes of death
im.cod.dat$cod_recode <- NA
im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == 
                        "#Congenital malformations, deformations and chromosomal abnormalities (Q00-Q99)"] <- "Congenital Malformations"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Congenital Malformations" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == 
                        "#Disorders related to short gestation and low birth weight, not elsewhere classified (P07)"] <- "Short Gestation/LBW"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Short Gestation/LBW" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Sudden infant death syndrome (R95)"] <- "SIDS"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "SIDS" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Newborn affected by maternal complications of pregnancy (P01)"] <- "Maternal Complication of Pregnancy"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Maternal Complication of Pregnancy" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Accidents (unintentional injuries) (V01-X59)"] <- "Accidents"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Accidents" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Newborn affected by complications of placenta, cord and membranes (P02)"] <- "Complications of Placenta, Cord, and Membranes"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Complications of Placenta, Cord, and Membranes" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Respiratory distress of newborn (P22)"] <- "Respiratory Distress"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Respiratory Distress" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Bacterial sepsis of newborn (P36)"] <- "Bacterial Sepsis"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Bacterial Sepsis" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Diseases of the circulatory system (I00-I99)"] <- "Diseases of the Circulatory System"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Diseases of the Circulatory System" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Neonatal hemorrhage (P50-P52,P54)"] <- "Neonatal Hemorrhage"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Neonatal Hemorrhage" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Intrauterine hypoxia and birth asphyxia (P20-P21)"] <- "Intrauterine hypoxia and birth asphyxia"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Intrauterine hypoxia and birth asphyxia" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Necrotizing enterocolitis of newborn (P77)"] <- "Necrotizing enterocolitis"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Necrotizing enterocolitis" & is.na(im.cod.dat$cod_recode) == F, ]

im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Atelectasis (P28.0-P28.1)"] <- "Atelectasis"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Atelectasis" & is.na(im.cod.dat$cod_recode) == F, ]


im.cod.dat$cod_recode[im.cod.dat$ICD10.130.Infants == "#Assault (homicide) (*U01,X85-Y09)"] <- "Assault"
table(im.cod.dat$cod_recode)
im.cod.dat[im.cod.dat$cod_recode == "Assault" & is.na(im.cod.dat$cod_recode) == F, ]

#only keep these rows of data
im.cod.dat2 <- im.cod.dat %>% filter(is.na(cod_recode) == F) %>%
  select(Race, Year, Deaths, Population, RatePer1000, cod_recode)

#calculate the infant mortality rate for the top fifteen causes of infant death
#and store this info in a new data frame called total14
total14 <- im.cod.dat2 %>% group_by(Race, Year) %>%
  summarise(Deaths.top14 = sum(Deaths))
  
```

```{r notes, echo = F, eval = F}
# 15 leading causes of death by ICD code for infants in 2014
# ref: https://www.cdc.gov/nchs/data/dvs/lcwk7_2014.pdf
#We only considered the top fourteen causes since the rate for the 15th causes was very low and unreliably coded in the earlier years

#Y Congenital malformations, deformations and chromosomal abnormalities (Q00-Q99)	64,181	47,147,656	136.1
#Y Disorders related to short gestation and low birth weight, not elsewhere classified (P07)	58,033	47,147,656	123.1
#Y Sudden infant death syndrome (R95)	29,760	47,147,656	63.1
#Y Newborn affected by maternal complications of pregnancy (P01)	20,990	47,147,656	44.5
#Y Accidents (unintentional injuries) (V01-X59)	15,215	47,147,656	32.3
#Y Newborn affected by complications of placenta, cord and membranes (P02)	13,411	47,147,656	28.4
#Y Respiratory distress of newborn (P22)	9,638	47,147,656	20.4
#Y Bacterial sepsis of newborn (P36)	8,888	47,147,656	18.9
#Y Diseases of the circulatory system (I00-I99)	7,119	47,147,656	15.1
#Y Neonatal hemorrhage (P50-P52,P54)	6,421	47,147,656	13.6
#Y Intrauterine hypoxia and birth asphyxia (P20-P21)	5,567	47,147,656	11.8
#Y Necrotizing enterocolitis of newborn (P77)	5,526	47,147,656	11.7
#Y Atelectasis (P28.0-P28.1)	4,681	47,147,656	9.9
#Y Assault (homicide) (*U01,X85-Y09)	4,027	47,147,656	8.5
#N Diarrhea and gastroenteritis of infectious origin (A09)
```

```{r setup_data, echo=F}
#read in a separate file, where the deaths are aggregated over all COD
#this dataset is stratified by gender, but the previous one was not
im.dat <-  read.csv("/Users/corinneriddell/Documents/repos/InfantMortality/Data/InfantMortalityData.csv",
                      header = T)

imdat.subset <- im.dat %>% 
  filter(Race %in% c("Black or African American", "White"),
         Hispanic_Origin == "Not Hispanic or Latino") %>%
  select(Race, Year, Gender, Deaths, Population, Crude_Rate) %>%
  mutate(Race = relevel(Race, ref = "White")) %>%
  mutate(Race = droplevels(Race)) %>%
  mutate(Race = plyr::revalue(Race, c("White"="White", "Black or African American"="Black")),
         Population = as.numeric(as.character(Population)),
         RatePer1000 = (Deaths/Population)*1000)

totals <- imdat.subset %>% group_by(Race, Year) %>% 
  summarise(Deaths.all = sum(Deaths), #this sums across gender
            Population = sum(Population)) 

#this dataset contains overall death counts ("Deaths.all") and the deaths from the top 14 causes ("Deaths.top14")
totals <- merge(totals, total14, by = c("Race", "Year"))

#calculate the rate of deaths from all other causes
totals <- totals %>% mutate(Deaths = Deaths.all - Deaths.top14,
                            RatePer1000 = (Deaths/Population)*1000,
                            cod_recode = "All other causes") %>%
                     select(Race, Year, Population, Deaths, RatePer1000, cod_recode) 

#add in the "All other causes" data to the COD-specific data
im.cod.dat3 <- rbind(im.cod.dat2, totals)  
im.cod.dat3$cod_recode <- factor(im.cod.dat3$cod_recode)

#order the factor variable 
im.cod.dat3$cod_recode <- reorder(im.cod.dat3$cod_recode, im.cod.dat3$RatePer1000, max)
im.cod.dat3$cod_recode2 <- factor(im.cod.dat3$cod_recode, levels = rev(levels(im.cod.dat3$cod_recode)))
```

```{r reshape_long_to_wide, echo=F}
black2 <- subset(im.cod.dat3, Race == "Black") %>% 
  select(Year, RatePer1000, Population, Deaths, cod_recode, cod_recode2) %>%
  rename(Black_IMR = RatePer1000, Black.Pop = Population,
         Black.Deaths = Deaths)

white2 <- subset(im.cod.dat3, Race == "White") %>% 
  select(Year, RatePer1000, Population, Deaths, cod_recode, cod_recode2) %>%
  rename(White_IMR = RatePer1000, White.Pop = Population,
         White.Deaths = Deaths)

both.cod <- merge(black2, white2, by = c("Year", "cod_recode", "cod_recode2"))
```

```{r COD_trends_IMR}
ggplot(both.cod, aes(x = Year, y = Black_IMR)) + 
  geom_ribbon(aes(ymin= Black_IMR, ymax = White_IMR), col = "grey", alpha=0.2) +
  geom_line(aes(col = "Black")) + facet_wrap(~cod_recode2, scales = "free_y") +
  geom_line(aes(y = White_IMR, col = "White")) +   
  theme_minimal() + ylab("Cause-Specific Infant Mortality Rate (per 1,000)") +
  ggtitle("Trends in cause-specific infant mortality rates over time for black and whites, United States 1999-2015") +
  scale_color_manual(values = c("#b2182b", "#2166ac"), guide = guide_legend(title = ""))

ggplot(subset(both.cod, Year >= 2010), aes(x = Year, y = Black_IMR)) + 
  geom_ribbon(aes(ymin= Black_IMR, ymax = White_IMR), col = "grey", alpha=0.2) +
  geom_line(aes(col = "Black")) + facet_wrap(~cod_recode2, scales = "free_y") +
  geom_line(aes(y = White_IMR, col = "White")) +   
  theme_minimal() + ylab("Cause-Specific Infant Mortality Rate (per 1,000)") +
  ggtitle("Trends in cause-specific infant mortality rates over time for black and whites, United States 2010-2015") +
  scale_color_manual(values = c("#b2182b", "#2166ac"), guide = guide_legend(title = ""))
```

-Two years only
-Stacked with largest contributions at the bottom

```{r}
ggplot(arrange(subset(both.cod, Year >= 2014), desc(cod_recode)), aes(x = Year, y = Black_IMR, fill=cod_recode)) +
    geom_bar(stat='identity', col = "white") + theme_minimal() + ylab("Infant mortality rate (per 1000)") +
  ggtitle("Black")

ggplot(arrange(subset(both.cod, Year >= 2014), desc(cod_recode)), aes(x = Year, y = White_IMR, fill=cod_recode)) +
    geom_bar(stat='identity', col = "white") + theme_minimal() + ylab("Infant mortality rate (per 1000)") +
  ggtitle("White")
```

-Past five years
-Stacked with largest contributions at the top

```{r}
ggplot(arrange(subset(both.cod, Year >= 2010), cod_recode), aes(x = Year, y = Black_IMR, fill=cod_recode)) +
    geom_bar(stat='identity', col = "white") + theme_minimal() + ylab("Infant mortality rate (per 1000)") +
  ggtitle("Black")

ggplot(arrange(subset(both.cod, Year >= 2010), cod_recode), aes(x = Year, y = White_IMR, fill=cod_recode)) +
    geom_bar(stat='identity', col = "white") + theme_minimal() + ylab("Infant mortality rate (per 1000)") +
  ggtitle("White")
```

```{r COD_RiskRatio}
both.cod <- both.cod %>% mutate(RR = Black_IMR/White_IMR, RD = Black_IMR - White_IMR)

ggplot(both.cod, aes(x = Year, y = RR))  + 
  geom_line(lwd = 1.5) + #, aes(col = cod_recode)) + 
  ylab("Risk Ratio") + facet_wrap(~cod_recode2) +
  theme_minimal() + ggtitle("Trends in the relative risk of cause-specific infant mortality over time comparing blacks to whites, United States 1999-2015")

ggplot(subset(both.cod, Year>= 2010), aes(x = Year, y = RR))  + 
  geom_line(lwd = 1.5) + #, aes(col = cod_recode)) + 
  ylab("Risk Ratio") + facet_wrap(~cod_recode2) +
  theme_minimal() + ggtitle("Trends in the relative risk of cause-specific infant mortality over time comparing blacks to whites, United States 1999-2015")
```

```{r COD_RiskDifference}
ggplot(both.cod, aes(x = Year, y = RD))  + 
  geom_line(lwd = 1.5) + #, aes(col = cod_recode)) + 
  ylab("Risk Difference (excess deaths per 1,000)") + facet_wrap(~cod_recode2, scales = "free_y") +
  theme_minimal() + ggtitle("Trends in the excess risk of cause-specific infant mortality over time comparing blacks to whites, United States 1999-2015")

ggplot(subset(both.cod, Year>= 2010), aes(x = Year, y = RD))  + 
  geom_line(lwd = 1.5) + #, aes(col = cod_recode)) + 
  ylab("Risk Difference (excess deaths per 1,000)") + facet_wrap(~cod_recode2, scales = "free_y") +
  theme_minimal() + ggtitle("Trends in the excess risk of cause-specific infant mortality over time comparing blacks to whites, United States 1999-2015")
```

```{r}
black <- subset(imdat.subset, Race == "Black") %>% 
  select(Year, Gender, RatePer1000, Population, Deaths) %>%
  rename(Black_IMR = RatePer1000, Black.Pop = Population,
         Black.Deaths = Deaths)

white <- subset(imdat.subset, Race == "White") %>% 
  select(Year, Gender, RatePer1000, Population, Deaths) %>%
  rename(White_IMR = RatePer1000, White.Pop = Population,
         White.Deaths = Deaths)

both <- merge(black, white, by = c("Year", "Gender"))
both <- both %>% mutate(RR = Black_IMR/White_IMR, RD = Black_IMR - White_IMR,
                        SD_RD = (((Black.Deaths*(Black.Pop - Black.Deaths))/(Black.Pop^2*(Black.Pop-1))) + ((White.Deaths*(White.Pop - White.Deaths))/(White.Pop^2*(White.Pop-1))))^0.5,
         UCL_RD = RD + 1.96*SD_RD, LCL_RD = RD - 1.96*SD_RD, 
                  SD_lnRR = ((1/Black.Deaths)-(1/(Black.Pop-Black.Deaths))+(1/White.Deaths)-(1/(White.Pop-White.Deaths)))^0.5,
         UCL_RR = exp(log(RR) + 1.96*SD_lnRR), LCL_RR = exp(log(RR) - 1.96*SD_lnRR))

#formulas for the standard errors of the RR and RD from Modern Epidemiology, Third Edition, Ch.14, pg 249 by Rothman, Greenland, and Lash
#my CIs for the RD are not equal to those produced by the epiR package
#not sure why this is.
aggregate.sex <- both %>% group_by(Year) %>%
  summarise(Black.Pop = sum(Black.Pop), White.Pop = sum(White.Pop),
            Black.Deaths = sum(Black.Deaths), White.Deaths = sum(White.Deaths)) %>%
  mutate(Black_IMR = (Black.Deaths/Black.Pop)*1000,
         White_IMR = (White.Deaths/White.Pop)*1000) %>%
  mutate(RR = Black_IMR/White_IMR, RD = Black_IMR - White_IMR,
         SD_RD = (((Black.Deaths*(Black.Pop - Black.Deaths))/(Black.Pop^2*(Black.Pop-1))) + ((White.Deaths*(White.Pop - White.Deaths))/(White.Pop^2*(White.Pop-1))))^0.5,
         UCL_RD = RD + 1.96*SD_RD, LCL_RD = RD - 1.96*SD_RD,
         SD_lnRR = ((1/Black.Deaths)-(1/(Black.Pop-Black.Deaths))+(1/White.Deaths)-(1/(White.Pop-White.Deaths)))^0.5,
         UCL_RR = exp(log(RR) + 1.96*SD_lnRR), LCL_RR = exp(log(RR) - 1.96*SD_lnRR),
         Gender = "Both Genders")

both2 <- rbind(both, aggregate.sex)
```

```{r imr_plot, echo=F, fig.width=9, fig.height=3}
first <- ggplot(subset(both2, Gender == "Both Genders" & Year >= 2010), aes(x = Year, y = Black_IMR))  + 
  geom_ribbon(aes(ymin= Black_IMR, ymax = White_IMR), col = "grey", alpha=0.2) +
  geom_line(aes(col = "Black"), lwd = 1.5) + 
  geom_line(aes(y = White_IMR, col = "White"), lwd = 1.5) +
  ylab("IMR (per 1,000)") +
  scale_color_manual(values = c("#b2182b", "#2166ac"), guide = guide_legend(title = "")) +
  theme_minimal() + theme(legend.position = "top") + ggtitle("a) Infant Mortality Rates (IMR)")

ggplot(both2, aes(x = Year, y = Black_IMR))  + 
  geom_ribbon(aes(ymin= Black_IMR, ymax = White_IMR), col = "grey", alpha=0.2) +
  geom_line(aes(col = "Black"), lwd = 1.5) + 
  geom_line(aes(y = White_IMR, col = "White"), lwd = 1.5) +
  facet_wrap(~ Gender) + ylab("Infant Mortality Rate (per 1,000)") +
  scale_color_manual(values = c("#b2182b", "#2166ac")) +
  theme_minimal()
```

```{r RR_plot, echo=F, fig.width=8, fig.height=2}
second <- ggplot(subset(both2, Gender == "Both Genders" & Year >= 2010), aes(x = Year, y = RR))  + 
  geom_line(lwd = 1.5) + 
  ylab("Risk Ratio") +
  theme_minimal() + ggtitle("b) Relative risk")

ggplot(both2, aes(x = Year, y = RR))  + 
  geom_line(lwd = 1.5) + 
  facet_wrap(~ Gender) + ylab("Relative risk") +
  theme_minimal()
```

```{r}
table.2by2 <-rbind(c(both2$Black.Deaths[1], both2$Black.Pop[1] - both2$Black.Deaths[1]), 
                   c(both2$White.Deaths[1], both2$White.Pop[1] - both2$White.Deaths[1])) 
epi.2by2(table.2by2, method = "cohort.count", conf.level = 0.95, units = 1000)
```


```{r RD_plot, echo=F, fig.width=8, fig.height=2}
third <- ggplot(subset(both2, Gender == "Both Genders" & Year >= 2010), aes(x = Year, y = RD))  + 
  geom_line(lwd = 1.5) + 
  ylab("Excess risk (per 1,000)") +
  theme_minimal() + ggtitle("c) Excess risk")

ggplot(both2, aes(x = Year, y = RD))  + 
  geom_line(lwd = 1.5) + 
  facet_wrap(~ Gender) +
  ylab("Excess risk") +
  theme_minimal()
```

```{r all_three, fig.width=8, fig.height=8}
grid.arrange(first, second, third, nrow = 3, heights = c(0.5, 0.25, 0.25))
```

Infant mortality for whites decreased from 5.01 deaths per 1,000 in 2014 to 4.95 deaths per 1,000 in 2015. At the same time, infant mortality among black infants increased from 11.01 to 11.23 per 1,000 over the same time period. While the absolute size of this increase among blacks may seem small (~ 0.22 additional deaths per 1,000 infants), it means that an additional 1XX black infants died of those 5XX,XXX born in 2015 (compared to the 2014 level).

The risk ratio increased to 2.27 from 2.19, implying 2.27 times more infant deaths among blacks compared with whites.

The risk difference, an absolute measure of risk was 6.00/1000 in 2014 and increased to 6.29/1000 in 2015. 
Number of black infant deaths in 2015: 6907 for 614972 population
If blacks had rates of whites in 2015: 3041 for 614972 population
--> 6907/3041 = 2.27, 6907-3041 = 3866 additional deaths per 614972 --> X per 1000







```{r more_plots, echo=F, eval=F}
a <- ggplotly(
ggplot(both, aes(y = RR, x = Year)) + geom_line() +
  theme_minimal() + facet_wrap(~ Gender) + ylab("")) %>%
  layout(yaxis = list(title = "Risk Ratio"))

b <- ggplotly(
ggplot(both, aes(y = RD, x = Year)) + geom_line() +
  theme_minimal() + facet_wrap(~ Gender) + ylab("")) %>%
  layout(yaxis = list(title = "Risk Difference"))

c <- ggplotly( 
ggplot(both, aes(y = Black_IMR, x = Year)) +
  geom_ribbon(aes(ymin = Black_IMR, ymax = White_IMR), col = "grey", alpha = 0.2) +
  facet_wrap(~ Gender) + 
  geom_line(col = "#b2182b") + 
  geom_line(aes(y = White_IMR), col = "#2166ac") +
  theme_minimal() + ylab("")) %>%
  layout(yaxis = list(title = "Infant Mortality Rate"))

subplot(c, a, b, nrows = 3, titleY = T)

ggplot(imdat.subset, aes(x = Year, y = RatePer1000)) + 
  geom_line(aes(col = Race), lwd = 2) +
  theme_minimal() + facet_wrap(~Gender) +
  ylab("Infant Mortality Rate (per 1,000)")

```

